#Gizzard Shad Pee

## A model of gizzard shad nutrient recycling

Readings:

@Vanni2002

```{r, echo=FALSE, warning=FALSE, fig.cap="*A simple example of the role animals can play in regulating nutrient cycling.*"}
# a digraph is a directed graph
# the model is written in the "dot" language
shad <-
"digraph{
  rankdir = LR;
{
node[shape='box']; T;G;S;
node[shape='plain']; W;O;
}
W[label='Watershed']
O[label='Downstream']
T[label='Total P']
G[label='Gizzard\nShad']
S[label='Sediments']
{rank=max; O;}
{rank=min; W;}
{rank=same; T;S;}

W -> T[label='inflow'];
T -> S[label='sedimentation                 '];
S -> T[label='                microbes'];
S -> G[label='ingestion'];
G -> T[label='translocation']; 
T -> O[label='outflow']; 
}"

grViz(shad)
grViz(shad) %>%
  export_svg %>% charToRaw %>% rsvg_pdf("shad.pdf")
```

## The model
Constants

* Gizzard shad excretion rate ($0.97\,\mu g$P $\,$L$^{-1}\,$d$^{-1}$)
* Inflow from the watershed  ($1.23\,\mu g$ P$\,$L$^{-1}\,$d${-1}$)
* Microbial release from the sediments ($0.12\,\mu g$P$\,$L$^{-1}\,$d$^{-1}$)

Primary production, PR, is a linear function of total water column phosphorus, $P_T$, 
$$ \mathrm{PR} = aP_T-b = 8.36 P_T - 115.9 $$
Sedimentation from the water column to the benthos is a ratio of the primary production and the export ratio, scaled by the P:C ratio (0.018, by mass). The export ratio, $E$, is itself a function of primary production,
$$E = c - d\cdot \mathrm{PR} = 0.459 - 0.000163\,\mathrm{PR} $$

Sedimentation, $S$, is thus,
$$S = \mathrm{PR}\cdot E \cdot 0.018$$
Sedimentation rate as a function of total phosporus in the water column (and various constants).
```{r, echo=TRUE}
translocation = 0.97
inflow=1.23
release=0.12
(inputs <- translocation+inflow+release)

a=8.36
b=115.9
c=0.459
d=0.000163

# P:C ratio
r = 0.018

# x is total phosphorus

## Sedimentation alone
{
par(mar=c(5,5,2,2))
curve( ( -a^2*d*x^2 + (a*c + 2*a*b*d)*x - (b*c - b^2*d) ) * r, 25, 125, 
       ylab=expression( "Sedimentation (" * mu *"g" * "P" * "L"^-1*"d"^-1*")"), 
       xlab = expression( "Total P (" * italic(mu) *"g" * "P" * "L"^-1*"d"^-1*")") )
}

## Inputs minus sedimentation
{
  par(mar=c(5,5,2,2))
  curve( inputs - ( -a^2*d*x^2 + (a*c + 2*a*b*d)*x - (b*c - b^2*d) ) , 25, 125, 
       ylab=expression( "Sedimentation (" * mu *"g" * "P" * "L"^-1*"d"^-1*")"), 
       xlab = expression( "Total P (" * italic(mu) *"g" * "P" * "L"^-1*"d"^-1*")") )
  abline(h=0, lty=3)
}
x=52
( -a^2*d*x^2 + (a*c + 2*a*b*d)*x - (b*c - b^2*d) ) * r
```


```{r shadmodel}

params <-c(translocation = 0.97, inflow=1.23, release=0.12, P.C=0.018)

y0 <- c(TP=40)

acton <- function(t,y,p) {

with( as.list(c(y,p)), {

 PPR = 8.36 * TP - 115.9
 ER = 0.459 - 0.000163 * PPR 
 sedimentation = PPR * ER * P.C
 
 dT.dt <- inflow + release + translocation - sedimentation 
 
 lR = 1.0077 * log10(TP) + 0.7206
 R = 10^(lR)
 return( list(c(dT.dt), PPR = PPR, Regen = R) )
} 
)

  
}

y0 <- c(TP=40)
days <- 0:200
sim.shad <- ode(y=y0, times= days, func=acton, parms=params)
plot(sim.shad)
```
